\documentclass{bioinfo}
\copyrightyear{2015}
\pubyear{2015}
\usepackage{natbib}
%\bibliographystyle{apalike}
\bibliographystyle{natbib.bst}
\newcommand{\mytitle}[2]{\title[#1]{#2}}
\newcommand{\myauthor}[2]{\author[#1]{#2}}
\newcommand{\myaddress}[1]{\address{#1}}
\newcommand{\myfootnote}[1]{\footnote{#1}}
\newcommand{\posttitle}[2]{}
\usepackage{xspace}
\newcommand{\ps}{phylesystem\xspace}
\newcommand{\otol}{Open Tree of Life\xspace}
\newcommand{\nexson}{otNexSON\xspace}
\newcommand{\js}{JavaScript\xspace}
\newcommand{\nexml}{NeXML\xspace}
\newcommand{\authorswaffil}{Emily Jane McTavish,$^{1}$
   Ruth Timme$^{2}$
}
\newcommand{\affil}{$^{1}$Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence KS, USA\\
$^{2}$ Center for Food Safety and Nutrition, Food and Drug Administration\\
}

\usepackage{hyperref}
\hypersetup{backref,  linkcolor=blue, citecolor=black, colorlinks=true, hyperindex=true}

\begin{document}
\firstpage{1}
\mytitle{Trer to Reads}{TreeToReads - a pipeline for simulating raw reads from phylogenies}

\myauthor{McTavish \textit{et~al}}{\authorswaffil}
\myaddress{\affil}
\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
\editor{Associate Editor: XXXXXXX}
\maketitle
\posttitle{\authorswaffil}{\affil}


\begin{abstract}
Using genome-wide SNP-based methods for tracking pathogens has become standard practice in academia and public health agencies across the world. 
There are multiple computational approaches available now that all perform a similar task: call variants by mapping short read data against a reference genome, filter these variants against set quality thresholds, then concatenate the variants, or SNPs, into a sequence matrix for downstream phylogenetic analysis. However, there are no existing methods to validate the accuracy of these approaches. 
We know there are multiple parameters that can affect whether or not the correct tree is recovered, such as a distant reference genome, low read coverage, low sequence diversity, etc. 
We present a simulatiion approach to generate realistic raw read data, allowing the user to vary  parameters of interest at each step i the simulation (tree shape, model of sequence evolution, distance of reference genome, etc.) so that the values of parameter can be explored individually.

\end{abstract}

\section{Introduction}
Genomics has revolutionized understanding of patterns and processes of evolution across a wide range of taxa.
In-depth investigation of very recent evolutionary radiations, in which only one or a handful of single nucleotide polymorphisms (SNPs) may differentiate lineages, 
is only possible in the light of genomic data, where the attempt is made to collect every single nucleotide.
In these examples where estimates of ancestry rely on a handful of data points, it is particularly important to ensure that analysis methods are validated and free from bias. 
Rigorous testing of the accuracy and robustness of novel approaches for making evolutionary estimates
from sequence data is necessary.

Standard phylogenetic methods, such as maximum likelihood assume that the data are gathered in an unbiased manor.
Despite the sheer quantities of genomic data, it is possible that biases in data processing of single nucleotide polymorphism (SNP) data from genomes
affect conclusions.
Importantly, these biases are not readily recognizable from standard measures of phylogenetic uncertainty, 
such as bootstrap proportions.
If biases are systematic, inappropriate methods will converge to an incorrect result with high confidence.

Such problems can be caused by various factors, 
including genotyping of single nucleotides which are polymorphic in a subset of the population \citep{mctavish_how_2015},
missing data cutoffs resulting in preferential inclusion of loci evolving at lower rates \citep{huang_unforeseen_2014},
and biases in read mapping due to choice of reference genome \citep{bertels_automated_2014} or different error rates amongst 
individual mapping algorithms (Pightling et al., 2014). .
Mis-estimation can become even more pronounced by interaction between these dataset biases and analysis choices;
for example using a model of evolution developed for sequence data 
on a panel of exclusively variable sites \citep{lewis_likelihood_2001}.
or choosing an inappropriate model of evolution \citep{sullivan_are_1997} has been shown to affect inferences.

The TreeToReads (TTR) simulation procedure described here captures realistic patterns of sequence variation across phylogenies in order to assess the robustness of evolutionary inferences from whole genome data to potential biases in the data collection and analysis pipeline.
TTR uses an observed or expected phylogeny, empirical base frequencies, a full general time reversible evolutionary rate matrix, and gamma distributed rate variation and an empirically informed spatial distribution of variant sites across the genome. 
TTR generates sequencing reads based on a reference or ``anchor'' genome, and a phylogeny.

\section{Methods}
The TreeToReads (TTR) pipeline generates short read data from genomes simulated along an input phylogeny.
TTR requires an input phylogeny with branch lengths and an ``anchor genome''. 
The anchor genome provides the genomic structure that will be used for the simulations.
One user-specified tip in the phylogeny will have this genome sequence.
Currently the anchor genome can only consist of a single contig, so multiple chromosomes must be simulated individually or concatenated within the anchor genome.
The user selects a number variable sites to be simulated on the genome.
This will be total number of sites that vary across all individuals, not the expected number of substitutions between any pair of individuals.
The user also provides parameters for the generalized time reversible (GTR) model with gamma distributed rate heterogeneity.
These parameters consist of rates of changes between all pairs of nucleotides, the expected frequency of nucleotides, 
and a shape parameter for the gamma distribution determining the relative rates of change at different sites. 
Values for these parameters may easily be estimated from empirical data using Paup \citep{swofford_paup:_2001} or other phylogenetic software.
The relative branch lengths are determined by the branch lengths of the input tree, but while the output branch lengths will be proportional to the input branch lengths, 
they will be scaled by the number of variable sites which are simulated.
The values of the input branch lengths will determine the probability that a single site is affected by multiple mutational events 
(e.g. multiple hit mutations resulting in homoplasy or three nucleotides observed at the same site across the sample).
The pipeline uses seq-gen \citep{rambaut_seq-gen:_1997} to simulate variable sites along the phylogeny.
The variable sites generated by seq-gen are placed on the anchor genome resulting in different genomic sequences at each tip.
The locations of mutations on the genome can be either drawn from a uniform distribution across the genome, 
or clustered according to parameters of a gamma distribution, which can be assigned in a configuration file.
This procedure creates a folder containing the simulated genomes for each tip in the tree.
Using these simulated genomes, TTR calls the read simulation software, ART \citep{huang_art:_2012}.
While currently TTR only supports automated generation of illumina paired end reads, 
the simulated genome files may be used with any ART parameter configuration.
Alternatively, these simulated genomes may be fed into pipelines for simulation of other types of next-generation sequencing data, 
for example such as SimRAD \citep{lepais_simrad:_2014}, if RAD-seq like simulated data are desired.
If ART is invoked to simulated read data, TTR will output a fastq folder containing folders with the names of each tip from the simulation tree. 
In each of these folders are the simulated reads  in .fastq format and .sam format file of the read alignments.
In addition the locations of mutations in the genome and a list of the base present in each tip at each variable site are provided.


\section{Case Study}
We used tree to reads to simulate short read data for an observed phylogeny of Salmonella Bareilly sequences associated with a 2010 slamonella outbreak.
Evolutionary sequence model parameters were estiamted from empirical SNP data. 
To reflect the observed clustering of mutations in this data, we used a clustering setting of 25\% mutations having a distance to the nearest mutation drawn from a gamma distribution with an average of 

We started with the observed tree (Fig 1a), and generated an read error profiles using the observed data, and generated short reads at different coverage levels,
and used these short reads to test the ability of the FDA's analysis pipeline to accurately call SNPs at different simluated sequence covereage levels (Fig 1b, Fig1c). 
The input and configuration files needed to run these simulations are included in the 'example' directory of the TreeToReads repository.


\section{Conclusions}
TreeToReads allows researchers to test the joint effects of choices in data set construction and subsequent phylogenetic analysis on data sets with a known true phylogenetic histories.
By anchoring simulated data in empirical estimates of phylogeny and mutation distributions, 
we can determine the robustness of phylogenetic analyses and model selection to variant ascertainment procedures.
The phylogenetic perspective in simulation testing of assembly and alignment tools has been lacking in genomic simulation software.
It is essential to include it because if there is phylogenetic signal in, for example, what proportion of reads can be mapped, due to distance from a reference genome \citep{bertels_automated_2014},
this can bias the final inference of topology.
This simulation framework will provide researchers with access to empirical evidence to discern among the numerous options available for read mapping, assembly, alignment and SNP phylogenetic approaches,
based on the specific characteristics of their question and dataset.

\bibliography{note}
\end{document}
